Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-06 tagless/tagless-final.tex
Conclusion

which shows the naive CPS transform of the original code. The result is
quite unoptimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with the |P| evaluator instead, our
running test gives the expected
\begin{code}
  P.repr = {P.st = Some <fun>;
            P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

We can easily modify the CBV CPS transform to pass a piece of state
along with the continuation. This lets us support mutable state. We
can also use reference variables of the metalanguage to emulate
reference variables in the object language.
\oleg{Should we have an example in the code?}
\jacques{In the code, yes.  Probably not in the paper, it is getting 
long as it is}

@@

We were dissatisfied with current approaches to DSL embedding: either
the performance was hurt by tagging/untagging and other interpretive
overhead, or the offered solutions required too much expressive power
from the host language's type system.  By shifting from an initial
algebra approach to one which leverages the natural co-algebraic structure
of the $\lambda$-calculus, we show that both of these issues can be 
simultaneously dealt with.

% We show how to do many things without tags, in a simple type system, 
% essentially by avoiding any (inductive) types.

\jacques{Should we mention that algebraic type tags give us a partly
intensional representation of terms.  This nicely allows pattern matching,
but as sits inside an extensional theory, ensuring that all traces 
of these tags disappear seems quite difficult without a very powerful
type system (enter GADTs).}

The overarching theme of our approach -- representation of object terms in 
\S\ref{ourapproach}, future-stage code in \S\ref{S:PE-lift} and static
values in \S\ref{S:PE-solution} and CPS values in \S\ref{S:CPS} --
demonstrates that pushing the computation towards the producers
\oleg{or, constructors. Ken said: functionalization} generally
requires simpler type systems (such as mere HM). The same idea appears
in many other areas (e.g., ammortized heap analysis, TFP/TLCA paper).
\oleg{I was told that speculation and informality is permitted in
  conclusions.}

separate compilation: separate
typechecking of terms without knowing which particular
hole they will be plugged into.

\subsection*{Acknowledgments}
We would like to thank Martin Sulzmann and Walid Taha 
for helpful discussions.

>>>>

which shows the naive CPS transformation of the original code. The result is
quite unoptimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with the |P| evaluator instead, our
running test gives the expected
\begin{code}
  P.repr = {P.st = Some <fun>;
            P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

We can modify the CBV CPS transformation to pass a piece of state along
with the continuation. This technique lets us support mutable state. We
can also add mutable references to the object language using mutable
references of the metalanguage.  The accompanying code illustrates these
extensions.

@@

We solve the problem of embedding a typed object language in a typed
metalanguage without using GADTs, dependent types, or a universal type.
Our family of interpreters include an evaluator, a compiler, a partial
evaluator, and CPS transformers.  It is patent that they never get stuck,
because we represent object types as metalanguage types.  This work
makes it safer and more efficient to embed domain\hyp specific languages
in practical metalanguages such as Haskell and ML\@.

Our main idea is to represent object programs not in an initial algebra
but using the existing coalgebraic structure of the $\lambda$-calculus.
More generally, to squeeze more invariants out of a type system as
simple as Hindley-Milner, we shift the burden of representation and
computation from consumers to producers: encoding object terms as calls
to metalanguage functions (\S\ref{ourapproach}); build dynamic terms
alongside static ones (\S\ref{S:PE-lift}); simulating type functions for
partial evaluation (\S\ref{S:PE-solution}) and CPS
transformation~(\S\ref{S:CPS}).  This shift also underlies fusion,
functionalization, and amortized complexity analysis.

When the metalanguage does provide higher-rank and higher-kind
polymorphism, we can type-check and compile an object term separately
from any interpreters it may be plugged into.

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.
